---
title: "Iris Data Quality and Exploratory Analysis" 
author: "Akash Bhardwaj" 
date: today 
format:
  html: 
    toc: true 
    toc-depth: 2 
    theme: cosmo 
execute: 
  echo: false
  warning: false 
  message: false
--- 

## Project objective 
 
This report demonstrates a reproducible workflow for inspecting, checking and 
visualising a structured biological dataset in R. 
 
The analysis uses the built-in `iris` dataset, which contains measurements from 
150 flowers across three species: setosa, versicolor and virginica. 
 
## Packages and setup 
 
```{r} 
library(tidyverse) 
library(janitor) 
library(skimr) 
library(here) 
 
iris_data <- iris %>% 
clean_names() 
``` 
 
## Data overview 
 
```{r} 
head(iris_data) 
str(iris_data) 
``` 
 
The dataset contains 150 observations and five variables: sepal length, sepal 
width, petal length, petal width and species. 
 
## Data-quality checks 
 
### Missing values 
 
```{r} 
missing_values <- iris_data %>% 
 summarise(across(everything(), ~ sum(is.na(.)))) %>% 
 pivot_longer( 
   cols = everything(), 
   names_to = "variable", 
   values_to = "missing_values" 
 ) 
 
missing_values 
``` 
 
No missing values were identified. 
 
### Duplicate rows 
 
```{r} 
duplicate_rows <- sum(duplicated(iris_data)) 
duplicate_rows 
``` 
 
One duplicated row was identified. It was retained because the available dataset 
documentation does not indicate whether it represents a data-entry duplication 
or a legitimate repeated observation. 
 
## Summary statistics 
 
```{r} 
summary_stats <- iris_data %>% 
 group_by(species) %>% 
 summarise( 
   number_of_samples = n(), 
   mean_sepal_length = mean(sepal_length), 
   sd_sepal_length = sd(sepal_length), 
   mean_petal_length = mean(petal_length), 
   sd_petal_length = sd(petal_length), 
   .groups = "drop" 
 ) 
 
summary_stats 
``` 
 
Each species has 50 observations. Mean sepal length and mean petal length are 
lowest in setosa and highest in virginica. 
 
## Visualisation 
 
```{r} 
sepal_plot <- ggplot( 
 iris_data, 
 aes(x = species, y = sepal_length, fill = species) 
) + 
 geom_boxplot(alpha = 0.8) + 
 labs( 
   title = "Sepal Length by Iris Species", 
   x = "Species", 
   y = "Sepal Length (cm)" 
 ) + 
 theme_minimal() + 
 theme(legend.position = "none") 
 
sepal_plot 
``` 
 
The boxplot shows differences in sepal length across the three species, with 
virginica showing the largest values and setosa the smallest. 

## Principal Component Analysis 
 
Principal component analysis (PCA) was used to reduce the four correlated flower 
measurements into components that capture the largest sources of variation. 
 
```{r} 
 
iris_numeric <- iris_data %>% 
 select(sepal_length, sepal_width, petal_length, petal_width) 
 
iris_scaled <- scale(iris_numeric) 
 
iris_pca <- prcomp(iris_scaled, center = TRUE, scale. = TRUE) 
 
pca_scores <- as.data.frame(iris_pca$x) %>% 
 mutate(species = iris_data$species) 
 
pca_plot <- ggplot( 
 pca_scores, 
 aes(x = PC1, y = PC2, colour = species) 
) + 
 geom_point(size = 3, alpha = 0.8) + 
 labs( 
   title = "Principal Component Analysis of Iris Measurements", 
   x = "Principal Component 1", 
   y = "Principal Component 2", 
   colour = "Species" 
 ) + 
 theme_minimal() 
 
pca_plot 
``` 
## Limitations 
 
This is a training dataset rather than a clinical or experimental biomedical 
dataset. It does not include metadata on experimental conditions, batch effects, 
measurement methods or participant-level characteristics. Therefore, the results 
demonstrate an analytical workflow but should not be interpreted as evidence for 
a biological or clinical conclusion. 
 
## Conclusion 
 
This project established a reproducible R workflow for data inspection, 
missing-value assessment, duplicate checking, descriptive statistics and 
visualisation. These steps form the foundation for quality-controlled biomedical 
data analysis. 

---

## 
